Exploratory: Optimal Resolution

### 2021-04-06 exploratory/exploratory.Rmd
pages upload

---
title: "Exploratory: Optimal Resolution"
author: "Courtney R Armour"
output: html_document
---
updated: `r Sys.Date()`
```{r setup_environment, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE)

@@

# Variables
models <- c("rf","glmnet","xgbTree","svmRadial","rpart2")
levels <- c("phylum","class","order","family","genus","otu","asv")
levels_names <- c("Phylum","Class","Order","Family","Genus","OTU","ASV")
pal <- park_palette("Everglades",length(models))
names(pal) <- models
pal2 <- park_palette("Everglades",length(models))
names(pal2) <- c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")

@@

kable(input_data) 
```

>>>>

---
title: "Exploratory: Optimal Resolution"
author: "Courtney R Armour"
output: html_document
---
updated: `r Sys.Date()`
```{r setup_environment, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE)
options(scipen=999)

@@

#given a number, adds correct commas
set_commas <- function(num){

}
```

```{r data_setup}
# Variables
models <- c("rf","glmnet","xgbTree","svmRadial","rpart2")
levels <- c("phylum","class","order","family","genus","otu","asv")
levels_names <- c("Phylum","Class","Order","Family","Genus","OTU","ASV")
pal <- park_palette("Everglades",length(models))
names(pal) <- models
pal2 <- park_palette("Everglades",length(models))
names(pal2) <- c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")

@@

Since there are some that believe ASVs generated by Mothur are not as good as ASVs generated by DADA2, I ran a comparison. I used DADA2 (v1.18.0) to generate ASVs and ran them through the same mikropml model pipeline. I ran dada(..., pool=T) to caputure more rare ASVs and subsampled to XX reads after processing.

Below is a summary table of the number of features at each level. Interestingly there are only `r formatC(input_data %>% filter(level=="dada2") %>% pull(n_features),big.mark = ",",format="f",digits=0)` ASVs identified with DADA2 compared to `r formatC(input_data %>% filter(level=="asv") %>% pull(n_features),big.mark = ",",format="f",digits=0)` with Mothur. After preprocessing there are just `r input_data %>% filter(level=="dada2") %>% pull(n_features_preproc)` ASVs, about half of what we find with Mothur.

```{r dada2_table}
kable(input_data) %>% 
  kable_styling(bootstrap_options = "striped")
```

```{r dada2_auc}

mothur_df <- full_df %>% 
  filter(level=="asv") %>% 
  mutate(type="Mothur")

mothur_auc_df <- auc_df %>% 
  filter(level=="asv") %>% 
  mutate(type="Mothur")

dada2_df <- NULL
dada2_auc_df <- NULL
for( m in models ){
  asv <- read_csv(paste0("../data/dada2/process/combined-dada2-",m,".csv")) %>%
    mutate(type="DADA2")
  dada2_df <- bind_rows(dada2_df,asv)

  m_auc.df <- select(asv,cv_metric_AUC,AUC,method,type) %>% 
    rename(test_AUC=AUC) %>% 
    pivot_longer(cols = c("cv_metric_AUC","test_AUC"),names_to="auc_type",values_to="AUC")

  dada2_auc_df <- bind_rows(dada2_auc_df,m_auc.df)
}

md_df <- bind_rows(mothur_df,dada2_df)

md_auc_df <- bind_rows(mothur_auc_df,dada2_auc_df)

```{r dada2_stats}
dada2_p_level <- read.csv("../data/dada2/analysis/dada2_pvalues_by_level.csv")
dada2_p_model <- read.csv("../data/dada2/analysis/dada2_pvalues_by_model.csv")
```

```{r dada2_plot_auc,out.width="50%"}
mod_md_df <- md_df %>% 
  select(cv_metric_AUC,AUC,method,level,type) %>% 
  rename(test_AUC=AUC,cv_AUC=cv_metric_AUC) %>% 
  pivot_longer(cols=c("cv_AUC","test_AUC"),names_to="auc_type",values_to="AUC") %>% 
  mutate(method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")))

mod_md_df %>% 
  filter(type=="DADA2") %>% 
  ggplot(aes(x=method, y=AUC, fill=auc_type)) + 
    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
    geom_boxplot(notch = TRUE) + 
    theme_bw() + xlab("") + ylab("AUROC") + 
    theme(legend.position ="top",
          axis.text.x = element_text(angle = 45,hjust = 1)) +
    ggtitle("DADA2 ASV Model Performance across Models")

mod_md_df %>% 
  ggplot(aes(x=type, y=AUC, fill=auc_type)) + 
    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
    geom_boxplot(notch = TRUE) + 
    facet_grid(.~method) +
    theme_bw() + xlab("") + ylab("AUROC") + 
    theme(legend.position ="top",
          axis.text.x = element_text(angle = 45,hjust = 1)) +
    ggtitle("DADA2 vs Mothur ASV Model Performance")

```{r dada2_plot_auc_wstat,out.width="50%"}

#subset to just test data
mod_md_auc_df <- md_auc_df %>% 
  filter(auc_type=="test_AUC") %>% 
  mutate(method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")))

#compare different models with DADA2 ASVs
sub_stats_level <- dada2_p_level %>% 
  mutate(level=factor(level,levels=c("asv","dada2"),
                            labels=c("Mothur","DADA2")),
         model1=factor(model1,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree")),
         model2=factor(model2,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
                       TRUE ~ "*")) %>% 
  mutate(y.position=seq.int(0.85,1.05,length.out = nrow(.))) %>% 
  rename(group1=model1,group2=model2) %>% 
  filter(sig=="NS")

mod_md_auc_df %>% 
  filter(type=="DADA2") %>% 
  ggplot(aes(x=method,y=AUC)) + 
    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
    geom_boxplot(aes(fill=method),notch = T) +
    scale_fill_manual(values = pal2) + 
    scale_y_continuous(limits=c(0.2,1.05)) +
    theme_bw() + xlab("") + ylab("AUROC") +
    theme(legend.position ="none",
          axis.text.x = element_text(angle = 45,hjust = 1),
          panel.grid.major.x = element_blank(),
          axis.text = element_text(size=14)) +
    ggtitle("DADA2") +
    stat_summary(fun = mean, color="white",geom = "point", shape = 18, size = 3,show.legend = FALSE) +
    stat_pvalue_manual(sub_stats_level,label="sig")

#compare Mother and DADA2 ASVs across models
sub_stats_model <- dada2_p_model %>% 
   mutate(level1=factor(level1,levels=c("asv","dada2"),
                            labels=c("Mothur","DADA2")),
          level2=factor(level2,levels=c("asv","dada2"),
                            labels=c("Mothur","DADA2")),
         model=factor(model,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
  mutate(sig=case_when(p_value >= 0.05 ~ "NS",
                       TRUE ~ "*")) %>% 
  #filter(sig=="NS") %>% 
  mutate(y.position=rep(0.85,length.out = nrow(.))) %>% 
  rename(group1=level1,group2=level2) %>% 
  mutate(p_value=case_when(p_value < 0.001 ~ round(p_value,digits=4),
                           TRUE ~ round(p_value,digits=2)))

mod_md_auc_df %>%
  rename(group=type,model=method) %>% 
  ggplot(aes(x=group,y=AUC,color=group)) +
    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
    geom_jitter() +
    stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
    #geom_boxplot(aes(fill=model),notch = T) +
    scale_y_continuous(limits=c(0.4,0.9)) +
    theme_bw() + xlab("") + ylab("AUROC") +
    theme(legend.position ="none",
          axis.text.x = element_text(angle = 45,hjust = 1),
          panel.grid.major.x = element_blank(),
          axis.text = element_text(size=14)) +
    ggtitle("DADA2 v Mothur ASVs") +
    facet_grid(~model) +
    stat_pvalue_manual(sub_stats_model,label="p_value")

```{r dada2_all, echo=F, message=F, out.width="100%",fig.height=3}
dada2_auc_df %>% 
  rename(level=type) %>% 
  bind_rows(auc_df) %>% 
  mutate(level=factor(level,levels=c("phylum","class","order","family","genus","otu","asv","DADA2"),
                            labels=c("Phylum","Class","Order","Family","Genus","OTU","ASV","DADA2")),
         method=factor(method,levels=c("rf","glmnet","xgbTree","svmRadial","rpart2"),
                       labels=c("Random Forest","Logistic Regression","XGBoost","SVM Radial","Decision Tree"))) %>% 
  filter(auc_type == "test_AUC") %>% 
  ggplot(aes(x=level,y=AUC,fill=method)) + 
    geom_hline(yintercept = 0.5,color="grey",lty="dashed") +
    geom_boxplot(notch = TRUE) + 
    theme_bw() + xlab("") + ylab("AUROC") + 
    theme(legend.position ="top",
          axis.text.x = element_text(angle = 45,hjust = 1)) +
    scale_fill_manual(values=pal2,name="Model")
